The motivation behind this Master's thesis is to experiment how to predict and bet a FIFA World Cup tournament. Since the project doesn't have any funding only freely accessible data sources are used. Only the outcome of the match is predicted not the final score.

Football, being one of the most popular sports in the world, attracts a lot of global attention especially during main tournaments like the FIFA World Cup. This global interest has attracted big investment banks like Goldman Sachs and UBS to predict the tournament outcome. Also, the research community has been active around this topic; mostly predicting the probabilities for winning the tournament or predicting the most probable course of the tournament \cite{groll2018prediction, groll2015prediction, leitner2010forecasting}. As far as I know, this opportunity to profit from these predictions has not been researched.

For many betting the outcome of a match is as important part of the experience as watching the game. As a result the size of the betting industry has grown to massive proportions\footnote{\url{https://www.caughtoffside.com/2018/07/05/world-cup-gambling-is-the-highest-its-ever-been/}}. This must be one the reason why the field of economics has researched this industry, especially the efficiency of the betting market. When markets are efficient all the available information is reflected into the prices and there is no chance to "beat the market". The efficient market hypothesis in betting has been discarded by Vlastakis et al. \cite{vlastakis2009efficient} and Kuypers \cite{kuypers2008}. Nevertheless, the general assumption is that the probabilities given by the betting industry are close to the true probabilities. For this reason, models are often benchmarked against the odds given by the betting industry or the odds are used as features. \cite{leitner2010forecasting} However, during major tournaments heavy one-sided bets, marketing purposes, and increased competition can force bookmakers' odds further from the true probabilities. These side effects might create fruitful opportunities which can be harvested with the right betting strategy.

From the investor's point of view profiting from the FIFA World Cup can be a risky opportunity. Some extreme patience is needed if after the first tournament investor is left with a negative profits and a promise that the strategy will be profitable in the long run. Waiting four years for the next tournament is a long time if you've already lost a part of your savings. Also, there is no guarantee that the strategy is able to keep up with the development that happens during the time between the tournaments. That's is why the betting strategy needs to be profitable in every tournament, not just in the long run.

Nowadays,  many games that simulate football have a vast collection of player attribute data. This data is freely available and to my knowledge has been successfully used once \cite{shin2014novel}. Could this player attribute data be aggregated and used to improve the prediction accuracy together with the data of historical matches?

Many different methods has been used to predict football matches. In the early work models used known probability distributions, like the Poisson distribution or the negative binomial distribution, to estimate the probability for a football match's score \cite{moroney1962facts, dixon1997}. In football, the favourite team has relatively low chance to win compared to many other sports \cite{anderson2013numbers}. These early methods required often a single parameter which reflected the team's ability as well as possible. This worked when teams were not equally skilled. When skill levels where similar more features where needed to differentiate the teams' abilities. Calculating this ability value out of multiple features was hard. More complex models where needed. Logistic regression models and random forest models have performed well with multiple features and in comparison to ranking methods \cite{groll2018prediction, 10.1007/978-3-319-29504-6_48}. Therefor in this thesis both linear models and random forest based models are used in the prediction.

This thesis will answer to these two questions: "Is it possible to bet on the FIFA World Cup matches profitably?" and "Is it possible to predict the tournaments more accurately than the betting market predicts with freely accessible data?". The idea is not to only monetize the abnormal odds but to use the average odds from the markets in a profitable way. This way "beating the bookie" in long run is possible. The main contributions of this thesis are:
\begin{enumerate}
    \item Reveal a way to predict a FIFA World Cup tournament profitably
    \item Demonstrate the value of a football video game data in prediction
    \item Point out how difficult it is to predict a draw correctly and how differently models estimate it compared to other outcomes
\end{enumerate}

The rest of the thesis is structured as follows: in Literature review, I go through the existing research on the topic. Next, in Section 3, I describe the dataset and how player attributes are aggregated to team-level attributes. Then, in Section 4, I describe the prediction models, betting strategies and the tournament simulation process. In Section 5, results from the simulation experiments are stated. Finally, I conclude and discuss future work in Section 6.